---
title: "R and databases (SQLite)"
output: html_document
---

```{r libraries, include=FALSE}
# load the required libraries, here DBI, RSQSLite, and tidyverse
library(DBI)
library(RSQLite)
library(tidyverse)
```


```{r db_connection, include=FALSE}
# establish a connection to the database, here an existing database but we could
# also create a new one from within R
streamChem <- dbConnect(RSQLite::SQLite(), "./stream_chemistry.sqlite")
```


```{r list_tables, include=FALSE}
# list all of the tables in our database
dbListTables(streamChem)
```


```{sql, connection=streamChem}

-- from our in-class database exercise:

-- We can group results based on data features that can be binned, 
-- this is particularly useful for aggregate functions. The query 
-- below extracts the minimum and maximum dissolved oxygen (DO) 
-- values for each sonde_event.
SELECT
  sonde_event_id,
  MIN(DO) AS min_DO,
  MAX(DO) AS max_DO
FROM sonde_data 
GROUP BY sonde_event_id;
```


### we can pass R values to our query

```{r create_variable}
thirdEvent <- 3
```


```{sql, connection=streamChem}
SELECT
  sonde_event_id,
  MIN(DO) AS min_DO,
  MAX(DO) AS max_DO
FROM sonde_data
WHERE sonde_event_id = ?thirdEvent
GROUP BY sonde_event_id;
```


## revisit the assignment in R

### run the SQL query you generated for the assignment in R; simply SQL that is run in R

```{sql connection = streamChem}
SELECT
  sonde_events.id,
  sonde_events.site_id AS sampleStream,
  sonde_events.site_id AS sampleStream,
  ROUND(AVG(atm_pressure.BP),1) AS averagePressure
FROM atm_pressure
JOIN sonde_events ON (sonde_events.id = atm_pressure.sonde_event_id)
GROUP BY sonde_events.id;
```


### or pull the data in to R and use R to generate the desired output

```{sql connection = streamChem, output.var = "atm_pressure"}
-- create a data frame of the atmospheric pressure data
SELECT *
FROM atm_pressure;
```


```{sql connection = streamChem, output.var = "sonde_events"}
-- create a data frame of the sonde events data
SELECT *
FROM sonde_events;
```


```{r}
# group the atmospheric data by the sonde_event_id, calculate the average, then
# join to the sonde events data to retrieve the site_id, just as we had done
# with SQL but here with R
atm_pressure %>% 
  group_by(sonde_event_id) %>% 
  summarise(averagePressure = round(mean(BP),1)) %>% 
  ungroup() %>% 
  inner_join(sonde_events, by = c("sonde_event_id" = "id")) %>% 
  select(id = sonde_event_id, sampleStream = site_id, averagePressure)
```

# add data to our database


```{sql connection=streamChem}
-- update our sonde_events table with a new event
INSERT INTO sonde_events (
  site_id, 
  date, 
  instrument_id, 
  K2_20
)
VALUES (
  "HWC", 
  "2003-10-10", 
  "blue",
  50.68
);
```


```{r}
# bring the sonde data from our 5th event into R
sonde_data_5 <- read_csv('../data/sonde_data_5.csv')
```


```{r}
# write the fifth event data to the database
dbWriteTable(streamChem, "sonde_data", sonde_data_5, append = TRUE)
```


--DELETE FROM sonde_events where id = 5;
--UPDATE sqlite_sequence SET seq = (SELECT MAX(id) FROM sonde_events) WHERE name="sonde_events";

--DELETE FROM sonde_data where sonde_event_id = 5;
--UPDATE sqlite_sequence SET seq = (SELECT MAX(id) FROM sonde_data) WHERE name="sonde_data";